---
  title: "RNAseq DESeq2 Time Course"
author: "Corey Hayford (modifed from Jack)"
date: "November 2018-January 2019"
output: pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/QuarantaLab/RNAseq_BP/")
```

This is a pipeline for differential analysis of RNASeq data from SKMEL5 sublines using DESeq2 statistical package. Three sublines: SC01 (*regressing*), SC07 (*stationary*) and SC10 (*expanding*) were analyzed for gene expression differences. In addition, time course changes in 8uM PLX4720 were also performed for each subline. Time points are: 0, 3d, 8d. The differential analysis will be performed based on the contrasts defined below. 
General steps for the analysis are:
  
###1. Read counts table: 
+ Could be read directly as a csv/txt file. 
+ Alignment and read counts could be done within R environment to create read counts table. 
1. Define working directory, load the required libraries. 
2. Get read counts table. 
Read the raw counts file processed by featureCounts. The fastq files were aligned with HiSat2, and the read counts were obtained using featureCounts of Rsubread packages.
```{r, echo=TRUE}
library(biomaRt)
d <- read.csv("featureCounts_matrix_all.csv", header=T, sep=",")

#Rename columns
cols <- c("ensembl_gene_id", "SC01_day0_rep1", "SC01_day0_rep2", "SC01_day0_rep3",
          "SC01_day3_rep1", "SC01_day3_rep2", "SC01_day3_rep3",
          "SC01_day8_rep1", "SC01_day8_rep2", "SC01_day8_rep3",
          "SC07_day0_rep1", "SC07_day0_rep2", "SC07_day0_rep3",
          "SC07_day3_rep1", "SC07_day3_rep2", "SC07_day3_rep3",
          "SC07_day8_rep1", "SC07_day8_rep2", "SC07_day8_rep3",
          "SC10_day0_rep1", "SC10_day0_rep2", "SC10_day0_rep3",
          "SC10_day3_rep1", "SC10_day3_rep2", "SC10_day3_rep3",
          "SC10_day8_rep1", "SC10_day8_rep2", "SC10_day8_rep3")
names(d) <- cols
ensembl <- useMart("ensembl")
mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
genes <- d$ensembl_gene_id
G_list <- getBM(attributes= c("ensembl_gene_id","hgnc_symbol"),
                filters= "ensembl_gene_id",
                values=genes,
                mart=mart)

GE_data <- merge(d, G_list, by = "ensembl_gene_id")
d <- GE_data[, -1]
d <- d[c(28, seq(1:27))]
rownames(d) <- make.names(d$hgnc_symbol, unique = T)
d <- d[, 2:28]

countdata <- d
# baseline <- c(1,2,3,10,11,12,19,20,21)
# treat3d  <- c(4,5,6,13,14,15,22,23,24)
# treat8d  <- c(7,8,9,16,17,18,25,26,27)
# # define the groups by subclones
# sc01 <- c(baseline[1:3], treat3d[1:3], treat8d[1:3])
# sc07 <- c(baseline[4:6], treat3d[4:6], treat8d[4:6])
# sc10 <- c(baseline[7:9], treat3d[7:9], treat8d[7:9])
# # Get the countdata specific to conditions: 
# # countdata <- countdata[,c(baseline)] 
# rownames(countdata) <- d[,"ensembl_gene_id"]
head(countdata)
nrow(countdata)
ncol(countdata)
```

### Identifying different ion channel gene lists
```{r}
countdata_histamine = countdata[c("HRH1","HRH2","HRH3","HRH4"),]
countdata_IPs = countdata[c("ITPR1", "ITPR2", "ITPR3"),]
countdata_IPrel = countdata[c("ITPR1", "ITPR2", "ITPR3", "PLCG1","DGKA", "DGKB", "DGKD", "DGKE", "DGKG", "DGKH", "DGKI", "DGKK", "DGKQ", "DGKZ", "PIK3CA", "PIK3CB", "PIK3CD", "PIK3CG", "PIK3C2A", "PIK3C2B", "PIK3C2G", "PIK3C3"),]
countdata_musc = countdata[c("CHRM1", "CHRM2", "CHRM3", "CHRM4", "CHRM5"),]
countdata_glut = countdata[c("GRM1", "GRM2", "GRM3", "GRM4", "GRM5",
                             "GRM6", "GRM7", "GRM8", "GRIA1", 
                             "GRIA2", "GRIA3", "GRIA4", "GRID1",
                             "GRID2", "GRIK1", "GRIK2", "GRIK3", 
                             "GRIK4", "GRIK5", "GRIN1", "GRIN2A",
                             "GRIN2B", "GRIN2C", "GRIN2D", "GRIN3A",
                             "GRIN3B"),]

### Plot IP# data
source("~/Documents/QuarantaLab/SummarySE.R")
IPs_match = colsplit(colnames(countdata_IPs), pattern = "_", names = c("Population", "Time", "Replicate"))
IPs_plot = cbind(IPs_match, t(countdata_IPs))
IPs_melt = melt(data = IPs_plot, id.vars = c("Population", "Time", "Replicate"), measure.vars = c("ITPR1", "ITPR2", "ITPR3"))
IPs_dat = summarySE(IPs_melt, measurevar = "value", groupvars = c("Population", "Time", "variable"))
IPs_dat$Time = as.numeric(gsub("[^[:digit:]]","",IPs_dat$Time))
IPs_dat_sub = subset(IPs_dat, variable %in% c("ITPR2","ITPR3"))
IPs_ggploted <- ggplot(IPs_dat_sub, aes(x=Time, y=value, group = interaction(variable, Population))) + geom_line(size=1.5, aes(color = Population, linetype = variable)) + geom_point(size = 1.5, aes(color = Population)) +
geom_errorbar(aes(ymin=value-sd, ymax=value+sd, color = Population), width=.2, size=1.5) +
theme_bw() + xlab("Time (days)") + ylab("Gene Counts") +
ggtitle("IP3 gene receptors") +
theme(legend.text = element_text(size = 10), legend.position = "right", 
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"), axis.text=element_text(size=12),
      legend.title = element_text(size=12,face="bold"),
      axis.title=element_text(size=12,face="bold"))

IPs_ggploted + ggsave("IP3R2_3_TimeSeriesRNAseq_clones.pdf", width = 6, height = 4)
```

###2. Convert counts table to DESeq2 object. 
Convert counts table to object for DESeq2 or any other analysis pipeline. This step will require to prepare data object in a form that is suitable for analysis in DESeq2 pipeline: we will need the following to proceed:
  
  + countdata: a table with the read/fragment counts. 
+ coldata: a table with information about the samples. 

Using the matrix of counts and the sample information table, we need to construct the DESeqDataSet object, for which we will use DESeqDataSetFromMatrix.....

#### 1. Define the samples and treatment conditions. 
```{r}
condition <- c("0", "3", "8")
treatment <- rep(condition, each=3) # Three biological replicates
unique(treatment)
cell <- c("SC01", "SC07","SC10") #sublines used for the analysis
cellName <- rep(cell, each=3)

coldata <- data.frame(cell=rep(cellName), treatment=rep(treatment, each=3))
group = factor(paste(coldata$cell, coldata$treatment, sep="."))
coldata$group = group
```

#### 2. construct the DESeqDataSet object from the matrix of counts and the sample information table. 
Described above are: countdata- raw counts, coldata: sample information table. 
```{r}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(countData = countdata,
                              colData = coldata,
                              design = ~ cell + treatment + cell:treatment)
dds
nrow(dds); ncol(dds)
```

###3. Exploratory analysis and visualization.
There are two separate steps in the workflow; the one which involves data transformations in order to visualize sample relationships and the second step involves statistical testing methods which requires the original raw counts. 

#### 1. Pre-filtering and normalization. 
Pre-filtering and normalization is required to remove lowly expressed genes. 

```{r}
dds2 <- dds[rowSums(counts(dds)) > 18, ] # remove rows with minimum of 2 read per condition
nrow(dds2)
save(dds2, file = "DDS_SC-1,7,10_cell-treat-int.RData")
# load("DDS_SC-1,7,10_cell-treat-int.RData")

```

#### 2. Visualize sample-to-sample distances. 
We could use Principal Component Analysis (PCA) to visualize relationships between samples. 
```{r}
library(dplyr)
rld <- rlog(dds2, blind = FALSE)
# save(rld, file = "RLD_SC-1,7,10_0,3,8d_20180701.RData")
# load("RLD_SC-1,7,10_0,3,8d_20180701.RData")
plotPCA(rld, intgroup = c("cell", "treatment"), ntop=5000)

## Use prcomp function
# Colored by cell line, shape by time point, lines connecting time
pca_DEseq <- prcomp(t(assay(rld)))
pca_DEseq_perc <- round(100*pca_DEseq$sdev^2/sum(pca_DEseq$sdev^2),1)
pca_DEseq_df <- data.frame(PC1 = pca_DEseq$x[,1], 
                           PC2 = pca_DEseq$x[,2], 
                           sample = colnames(assay(rld)),
                           cell.line = rep(c("SC01", "SC07", "SC10"), each = 9),
                           day = rep(c("Day0", "Day3", "Day8"), each = 3),
                           replicate = rep(c("Rep1", "Rep2", "Rep3"), times=9))

pca_DEseq_means <- ddply(pca_DEseq_df, .(cell.line, day), summarise, meanPC1 = mean(PC1), meanPC2 = mean(PC2))

plt_pca <- ggplot(pca_DEseq_df, aes(PC1,PC2, color = cell.line))+
  geom_point(aes(shape = day), size=2) +
  geom_path(data = pca_DEseq_means, 
            aes(x=meanPC1, y=meanPC2,
                color=cell.line), arrow = arrow(),
            size = 0.8) +
  labs(x=paste0("PC1 (",pca_DEseq_perc[1],"% variance)"), y=paste0("PC2 (",pca_DEseq_perc[2],"% variance)")) +
  theme_bw() + scale_color_manual(values = c("orange", "cyan", "purple")) +
  # ggtitle("PCA - Subclones in Time") +
  theme(legend.text = element_text(size = 12), 
        plot.title = element_text(size = 12, 
                                  hjust = 0.5), 
        axis.text=element_text(size=12),
        legend.title = element_text(size=12),
        legend.position = "right",
        axis.title=element_text(size=12),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())

plt_pca + ggsave("SKMEL5_sublines_timeSeriesRNA_rld_Fig.svg", width = 4, height = 3)

library(ggpubr)  
plt_pca_leg <- get_legend(plt_pca)
as_ggplot(plt_pca_leg) + ggsave("SKMEL5_sublines_timeSeriesRNA_rld_Figleg.svg")


```

```{r}
t0 <- subset(pca_DEseq_df, day == "Day0")[,c(1,2)]
t3 <- subset(pca_DEseq_df, day == "Day3")[,c(1,2)]
t8 <- subset(pca_DEseq_df, day == "Day8")[,c(1,2)]

t0_dist <- dist(cbind(t0$PC1, t0$PC2), diag = T, upper = T)
t0_m <- as.matrix(t0_dist)
dimnames(t0_m) <- list(1:length(t0$PC1),1:length(t0$PC2))
t0_xy <- t(combn(colnames(t0_m),2))
t0_df <- data.frame(t0_xy, dist=t0_m[t0_xy])
t0_df$Condition = "Day0"

t3_dist <- dist(cbind(t3$PC1, t3$PC2), diag = T, upper = T)
t3_m <- as.matrix(t3_dist)
dimnames(t3_m) <- list(1:length(t3$PC1),1:length(t3$PC2))
t3_xy <- t(combn(colnames(t3_m),2))
t3_df <- data.frame(t3_xy, dist=t3_m[t3_xy])
t3_df$Condition = "Day3"

t8_dist <- dist(cbind(t8$PC1, t8$PC2), diag = T, upper = T)
t8_m <- as.matrix(t8_dist)
dimnames(t8_m) <- list(1:length(t8$PC1),1:length(t8$PC2))
t8_xy <- t(combn(colnames(t8_m),2))
t8_df <- data.frame(t8_xy, dist=t8_m[t8_xy])
t8_df$Condition = "Day8"


SC01 <- subset(pca_DEseq_df, cell.line == "SC01")[,c(1,2)]
SC07 <- subset(pca_DEseq_df, cell.line == "SC07")[,c(1,2)]
SC10 <- subset(pca_DEseq_df, cell.line == "SC10")[,c(1,2)]

SC01_dist <- dist(cbind(SC01$PC1, SC01$PC2), diag = T, upper = T)
SC01_m <- as.matrix(SC01_dist)
dimnames(SC01_m) <- list(1:length(SC01$PC1),1:length(SC01$PC2))
SC01_xy <- t(combn(colnames(SC01_m),2))
SC01_df <- data.frame(SC01_xy, dist=SC01_m[SC01_xy])
SC01_df$Condition = "SC01"

SC07_dist <- dist(cbind(SC07$PC1, SC07$PC2), diag = T, upper = T)
SC07_m <- as.matrix(SC07_dist)
dimnames(SC07_m) <- list(1:length(SC07$PC1),1:length(SC07$PC2))
SC07_xy <- t(combn(colnames(SC07_m),2))
SC07_df <- data.frame(SC07_xy, dist=SC07_m[SC07_xy])
SC07_df$Condition = "SC07"

SC10_dist <- dist(cbind(SC10$PC1, SC10$PC2), diag = T, upper = T)
SC10_m <- as.matrix(SC10_dist)
dimnames(SC10_m) <- list(1:length(SC10$PC1),1:length(SC10$PC2))
SC10_xy <- t(combn(colnames(SC10_m),2))
SC10_df <- data.frame(SC10_xy, dist=SC10_m[SC10_xy])
SC10_df$Condition = "SC10"


SC01_t0 <- subset(pca_DEseq_df, cell.line == "SC01" & day == "Day0")[,c(1,2)]
SC07_t0 <- subset(pca_DEseq_df, cell.line == "SC07" & day == "Day0")[,c(1,2)]
SC10_t0 <- subset(pca_DEseq_df, cell.line == "SC10" & day == "Day0")[,c(1,2)]
SC01_t3 <- subset(pca_DEseq_df, cell.line == "SC01" & day == "Day3")[,c(1,2)]
SC07_t3 <- subset(pca_DEseq_df, cell.line == "SC07" & day == "Day3")[,c(1,2)]
SC10_t3 <- subset(pca_DEseq_df, cell.line == "SC10" & day == "Day3")[,c(1,2)]
SC01_t8 <- subset(pca_DEseq_df, cell.line == "SC01" & day == "Day8")[,c(1,2)]
SC07_t8 <- subset(pca_DEseq_df, cell.line == "SC07" & day == "Day8")[,c(1,2)]
SC10_t8 <- subset(pca_DEseq_df, cell.line == "SC10" & day == "Day8")[,c(1,2)]

SC01_t0_dist <- dist(cbind(SC01_t0$PC1, SC01_t0$PC2), diag = T, upper = T)
SC01_t0_m <- as.matrix(SC01_t0_dist)
dimnames(SC01_t0_m) <- list(1:length(SC01_t0$PC1),1:length(SC01_t0$PC2))
SC01_t0_xy <- t(combn(colnames(SC01_t0_m),2))
SC01_t0_df <- data.frame(SC01_t0_xy, dist=SC01_t0_m[SC01_t0_xy])
SC01_t0_df$Condition = "SC01_t0"

SC07_t0_dist <- dist(cbind(SC07_t0$PC1, SC07_t0$PC2), diag = T, upper = T)
SC07_t0_m <- as.matrix(SC07_t0_dist)
dimnames(SC07_t0_m) <- list(1:length(SC07_t0$PC1),1:length(SC07_t0$PC2))
SC07_t0_xy <- t(combn(colnames(SC07_t0_m),2))
SC07_t0_df <- data.frame(SC07_t0_xy, dist=SC07_t0_m[SC07_t0_xy])
SC07_t0_df$Condition = "SC07_t0"

SC10_t0_dist <- dist(cbind(SC10_t0$PC1, SC10_t0$PC2), diag = T, upper = T)
SC10_t0_m <- as.matrix(SC10_t0_dist)
dimnames(SC10_t0_m) <- list(1:length(SC10_t0$PC1),1:length(SC10_t0$PC2))
SC10_t0_xy <- t(combn(colnames(SC10_t0_m),2))
SC10_t0_df <- data.frame(SC10_t0_xy, dist=SC10_t0_m[SC10_t0_xy])
SC10_t0_df$Condition = "SC10_t0"

SC01_t3_dist <- dist(cbind(SC01_t3$PC1, SC01_t3$PC2), diag = T, upper = T)
SC01_t3_m <- as.matrix(SC01_t3_dist)
dimnames(SC01_t3_m) <- list(1:length(SC01_t3$PC1),1:length(SC01_t3$PC2))
SC01_t3_xy <- t(combn(colnames(SC01_t3_m),2))
SC01_t3_df <- data.frame(SC01_t3_xy, dist=SC01_t3_m[SC01_t3_xy])
SC01_t3_df$Condition = "SC01_t3"

SC07_t3_dist <- dist(cbind(SC07_t3$PC1, SC07_t3$PC2), diag = T, upper = T)
SC07_t3_m <- as.matrix(SC07_t3_dist)
dimnames(SC07_t3_m) <- list(1:length(SC07_t3$PC1),1:length(SC07_t3$PC2))
SC07_t3_xy <- t(combn(colnames(SC07_t3_m),2))
SC07_t3_df <- data.frame(SC07_t3_xy, dist=SC07_t3_m[SC07_t3_xy])
SC07_t3_df$Condition = "SC07_t3"

SC10_t3_dist <- dist(cbind(SC10_t3$PC1, SC10_t3$PC2), diag = T, upper = T)
SC10_t3_m <- as.matrix(SC10_t3_dist)
dimnames(SC10_t3_m) <- list(1:length(SC10_t3$PC1),1:length(SC10_t3$PC2))
SC10_t3_xy <- t(combn(colnames(SC10_t3_m),2))
SC10_t3_df <- data.frame(SC10_t3_xy, dist=SC10_t3_m[SC10_t3_xy])
SC10_t3_df$Condition = "SC10_t3"

SC01_t8_dist <- dist(cbind(SC01_t8$PC1, SC01_t8$PC2), diag = T, upper = T)
SC01_t8_m <- as.matrix(SC01_t8_dist)
dimnames(SC01_t8_m) <- list(1:length(SC01_t8$PC1),1:length(SC01_t8$PC2))
SC01_t8_xy <- t(combn(colnames(SC01_t8_m),2))
SC01_t8_df <- data.frame(SC01_t8_xy, dist=SC01_t8_m[SC01_t8_xy])
SC01_t8_df$Condition = "SC01_t8"

SC07_t8_dist <- dist(cbind(SC07_t8$PC1, SC07_t8$PC2), diag = T, upper = T)
SC07_t8_m <- as.matrix(SC07_t8_dist)
dimnames(SC07_t8_m) <- list(1:length(SC07_t8$PC1),1:length(SC07_t8$PC2))
SC07_t8_xy <- t(combn(colnames(SC07_t8_m),2))
SC07_t8_df <- data.frame(SC07_t8_xy, dist=SC07_t8_m[SC07_t8_xy])
SC07_t8_df$Condition = "SC07_t8"

SC10_t8_dist <- dist(cbind(SC10_t8$PC1, SC10_t8$PC2), diag = T, upper = T)
SC10_t8_m <- as.matrix(SC10_t8_dist)
dimnames(SC10_t8_m) <- list(1:length(SC10_t8$PC1),1:length(SC10_t8$PC2))
SC10_t8_xy <- t(combn(colnames(SC10_t8_m),2))
SC10_t8_df <- data.frame(SC10_t8_xy, dist=SC10_t8_m[SC10_t8_xy])
SC10_t8_df$Condition = "SC10_t8"

test_df <- rbind(t0_df, t3_df, t8_df, SC01_df, SC07_df, SC10_df)
test_df_baseline <- rbind(SC01_t0_df, SC07_t0_df, SC10_t0_df,
                          SC01_t3_df, SC07_t3_df, SC10_t3_df,
                          SC01_t8_df, SC07_t8_df, SC10_t8_df)
test_df_baseline$Condition <- "Baseline"
test_df_all <- rbind(test_df, test_df_baseline)
ggplot(test_df_all, aes(dist, color=Condition)) + stat_ecdf()

library(transport)
EMD_df <- data.frame(Condition = c("SC01", "SC07", "SC10",
                                   "Day0", "Day3", "Day8"),
                     EMD = c(wasserstein1d(test_df_baseline$dist, SC01_df$dist),
                             wasserstein1d(test_df_baseline$dist, SC07_df$dist),
                             wasserstein1d(test_df_baseline$dist, SC10_df$dist),
                             wasserstein1d(test_df_baseline$dist, t0_df$dist),
                             wasserstein1d(test_df_baseline$dist, t3_df$dist),
                             wasserstein1d(test_df_baseline$dist, t8_df$dist)))

ggplot(EMD_df, aes(x=Condition, y=EMD, group = Condition, fill = Condition)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(name = "", values = c("grey90", "grey60", "grey30",
                                          "orange", "cyan", "purple")) +
  theme_classic() + ylab("EMD from Baseline") +
  theme(axis.text = element_text(size = 12, color = "black"),
        legend.position = "none", axis.title = element_text(size = 12, color = "black")) +
  ggsave("EMD_bulkRNAseq_byCondition.pdf", width = 4, height = 3)
  
```



```{r}
t0 <- subset(pca_DEseq_df, day == "Day0")[,c(1,2)]
t3 <- subset(pca_DEseq_df, day == "Day3")[,c(1,2)]
t8 <- subset(pca_DEseq_df, day == "Day8")[,c(1,2)]

t0$PC1dist <- (t0$PC1-colMeans(t0)[1])**2
t0$PC2dist <- (t0$PC2-colMeans(t0)[2])**2
t0$dist <- sqrt(t0$PC1dist + t0$PC2dist)

t3$PC1dist <- (t3$PC1-colMeans(t3)[1])**2
t3$PC2dist <- (t3$PC2-colMeans(t3)[2])**2
t3$dist <- sqrt(t3$PC1dist + t3$PC2dist)

t8$PC1dist <- (t8$PC1-colMeans(t8)[1])**2
t8$PC2dist <- (t8$PC2-colMeans(t8)[2])**2
t8$dist <- sqrt(t8$PC1dist + t8$PC2dist)

dist_mat <- data.frame(Sample = rep(c("Day0", "Day3", "Day8"), each = 9),
                       Distance = c(t0$dist, t3$dist, t8$dist))

dist_mat_melt <- melt(dist_mat, id.vars = "Sample", measure.vars = "Distance")
dist_mat_melt.summary <- aggregate(dist_mat_melt, by=list(dist_mat_melt$Sample), FUN=mean)[c(1,4)]
names(dist_mat_melt.summary) <- c("Sample", "Distance")

ggplot(dist_mat, aes(x=Sample, y=Distance, group = Sample,
                     color = Sample)) +
  geom_jitter(width = 0.2) + theme_bw() +
  geom_crossbar(data = dist_mat_melt.summary, aes(ymin=Distance, ymax=Distance, color = Sample),
                size=0.5, width = 0.5) +
  ylim(0,60) + scale_color_manual(values = c("black", "grey40", "gray80")) +
  labs(x = "Condition", y = "Distance from Mean") +
  theme(axis.text.y = element_text(size = 14),
        legend.position = "none", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=14),
        legend.title = element_text(size=14), axis.title=element_text(size=14),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  ggsave("SKMEL5_clones_distFromMean_byTP.svg", width = 2.5, height = 5)



```

### 4. Differential Expression Analysis. 
Always make sure to use the unnormalized raw counts for this. We will use DESeq function to perform differential analysis between samples; Unless specified, the analysis is between the last group and the first group. Different comparison can be done using 'contrast' argument. Steps involved underneath:
  
  1. estimation of size factors (controls for differences in sequencing depth of the samples)
2. estimation of dispersion values for each gene,
3. fitting a generalized linear model

#### 1. Running the differential expression pipeline. 
```{r, cache=TRUE}
design(dds2) = ~ cell + treatment + cell:treatment
dds <- DESeq(dds2, test = "LRT", reduced = ~ cell + treatment)
save(dds, file = "DESeq_SC1,7,10_Timecourse_LRT.RData")
# load("DESeq_SC1,7,10_Timecourse_LRT.RData")
# dds
```

#### 2. Building the results table. 
By default, results will extract the estimated log2 fold changes and p values for the last variable in the design formula. If there are more than 2 levels for this variable, results will extract the results table for a comparison of the last level over the first level. 
```{r}
# Esimate the differences between groups by: # a) Lowering the FDR (padj) or (b) raise the log2 fold change.

resultsNames(dds)
# alpha = FDR adjusted p value cutoff
res <- results(dds, alpha = 0.001)
summary(res)
resOrdered <- res[order(res$pvalue),]
rdata = as.data.frame(res)
```

```{r}
library(dplyr)
res_0to8d <- results(dds, name="treatment_8_vs_0", cooksCutoff = 0.99, 
                     independentFiltering = TRUE, alpha = 0.05, pAdjustMethod = "BH")
summary(res_0to8d)
# order results table by the smallest adjusted p value:
res_0to8d <- res_0to8d[order(res_0to8d$padj),]
results_0to8d <- as.data.frame(res_0to8d)

results_0to8d <- mutate(results_0to8d, sig=ifelse(results_0to8d$padj<0.05 & results_0to8d$log2FoldChange > 2, "Upregulated", ifelse(results_0to8d$padj<0.05 & results_0to8d$log2FoldChange < -2, "Downregulated", "Not Significant")))

row.names(results_0to8d) <- row.names(res_0to8d)


head(results_0to8d)
DEgenes_0to8d <- results_0to8d[which(abs(results_0to8d$log2FoldChange) > log2(1.5) & results_0to8d$padj < 0.05),]
```

```{r}
library(ggplot2)
library(ggrepel)
volcano <- ggplot(results_0to8d, aes(log2FoldChange, -log10(pvalue))) +
  geom_point(aes(col = sig)) + theme_bw() +
  scale_color_manual(values = c("red", "grey", "green3")) +
  # ggtitle("Volcano Plot of Untreated vs Idling") +
  labs(x="log2(Fold Change)", y="Log(Odds Ratio)") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.text = element_text(size = 12),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=12),
        legend.title = element_text(size=12), 
        axis.title=element_text(size=12),
        legend.position = "none")

volcano
# volcano + ggrepel::geom_text_repel(data=results_0to8d[1:10, ], 
#                                    ggplot2::aes(label=rownames(results_0to8d[1:10, ])))
# save(results_0to8d, file="untreatedIdling_DEA.RData")
```

# Generating Ion Channel Specific Gene Dataframes
```{r}
test <- assay(dds)
types <- c("ATP", "TRP", "GABR", "CRACR", "SLC", "KCN", "CACN", "GRI", "ABC", "SCN", "TRP", "RIC3", "CHRND", "RYR")
samples <- c("SC01_day0", "SC01_day3", "SC01_day8", "SC07_day0", "SC07_day3", "SC07_day8", "SC10_day0", "SC10_day3", "SC10_day8")
test <- test[grep(paste(types, collapse="|"), rownames(test)),]
test1 <- sapply(samples, function(x) rowMeans(test[, grep(x, colnames(test))]))
rownames(test1) <- rownames(test)
test1 <- test1[order(rownames(test1)),]
test2 <- as.data.frame(test1)
test2["l2FC_SC01_0to8"] <- log2(test2["SC01_day8"]/test2["SC01_day0"])
test2["l2FC_SC07_0to8"] <- log2(test2["SC07_day8"]/test2["SC07_day0"])
test2["l2FC_SC10_0to8"] <- log2(test2["SC10_day8"]/test2["SC10_day0"])
test3 <- subset(test2, l2FC_SC01_0to8 > 1 & l2FC_SC07_0to8 > 1 & l2FC_SC10_0to8 > 1)
test4 <- subset(test2, l2FC_SC01_0to8 > 1 | l2FC_SC07_0to8 > 1 | l2FC_SC10_0to8 > 1)
# write.csv(x = test2, file = "all_ionChannel_Expression.csv")
# write.csv(x = test3, file = "allUpreg_ionChannel_Expression.csv")
# write.csv(x = test4, file = "atLeastOneUpreg_ionChannel_Expression.csv")
test5 <- log2(test4[, 1:9]+1)
pheatmap(test5, cluster_cols = F, cluster_rows = F)
test6 <- log2((test3[,1:9])+1)
pheatmap(test6, cluster_rows = F, cluster_cols = F)
test7 <- test5[rowSums(test5)>30,]
pheatmap(test7, cluster_rows = F, cluster_cols = F)
```

```{r}
library(org.Hs.eg.db)
library(clusterProfiler)
# load(file="untreatedIdling_DEA.RData")

OrgDB <- org.Hs.eg.db
upreg_genes <- subset(results_0to8d, padj<0.05 & log2FoldChange>2)
downreg_genes <-subset(results_0to8d, padj<0.05 & log2FoldChange<(-2))

geneList_up <- as.vector(upreg_genes$log2FoldChange)
names(geneList_up) <- rownames(upreg_genes)
geneList_down <- as.vector(downreg_genes$log2FoldChange)
names(geneList_down) <- rownames(downreg_genes)

genes_up <- as.vector(rownames(upreg_genes))
genes_down <- as.vector(rownames(downreg_genes))
# names(geneList) <- rownames(results_0to8d)
genes_up_ENTREZID <- bitr(genes_up, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = OrgDB)$ENTREZID
genes_down_ENTREZID <- bitr(genes_down, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = OrgDB)$ENTREZID

# Group GO
ggo_up <- clusterProfiler::groupGO(gene     = genes_up_ENTREZID,
                                OrgDb    = OrgDB,
                                ont      = "BP",
                                level    = 3,
                                readable = TRUE)
ggo_up_df <- as.data.frame(ggo_up)
ggo_up_df <- ggo_up_df[order(-ggo_up_df$Count),] 

ggo_down <- clusterProfiler::groupGO(gene = genes_down_ENTREZID,
                                OrgDb    = OrgDB,
                                ont      = "BP",
                                level    = 3,
                                readable = TRUE)
# View(as.data.frame(ggo_down))

# GO over-representation test
ego_genesUp_BP <- clusterProfiler::enrichGO(gene  = genes_up_ENTREZID,
                                 OrgDb         = OrgDB,
                                 ont           = "BP",
                                 pAdjustMethod = "BH",
                                 pvalueCutoff  = 0.05,
                                 qvalueCutoff  = 0.05, 
                                 readable      = TRUE)
ego_genesUp_MF <- clusterProfiler::enrichGO(gene  = genes_up_ENTREZID,
                                 OrgDb         = OrgDB,
                                 ont           = "MF",
                                 pAdjustMethod = "BH",
                                 pvalueCutoff  = 0.05,
                                 qvalueCutoff  = 0.05, 
                                 readable      = TRUE)
ego_genesUp_CC <- clusterProfiler::enrichGO(gene  = genes_up_ENTREZID,
                                 OrgDb         = OrgDB,
                                 ont           = "CC",
                                 pAdjustMethod = "BH",
                                 pvalueCutoff  = 0.05,
                                 qvalueCutoff  = 0.05, 
                                 readable      = TRUE)

# View(as.data.frame(ego_genesUp))

ego_genesDown <- clusterProfiler::enrichGO(gene  = genes_down_ENTREZID,
                                 OrgDb         = OrgDB,
                                 ont           = "BP",
                                 pAdjustMethod = "BH",
                                 pvalueCutoff  = 0.05,
                                 qvalueCutoff  = 0.05, 
                                 readable      = TRUE)

# View(as.data.frame(ego_genesDown))

# kk_genesUp <- enrichKEGG(gene = genes_up_ENTREZID,
#                    organism = 'hsa',
#                   pvalueCutoff = 0.05)
# View(as.data.frame(kk_genesUp))
# 
# kk_genesDown <- enrichKEGG(gene = genes_down_ENTREZID,
#                    organism = 'hsa',
#                   pvalueCutoff = 0.05)
# View(as.data.frame(kk_genesDown))

# ego_GSEA_up <- gseGO(geneList = geneList_up,
#               OrgDb        = OrgDB,
#               ont          = "BP",
#               nPerm        = 1000,
#               minGSSize    = 100,
#               maxGSSize    = 500,
#               pvalueCutoff = 0.05,
#               verbose      = FALSE)

# barplot(ggo_up, order=T)
# barplot(ggo_down)
dotplot(ego_genesUp_MF) + ggtitle("GO Over-representation Upregulated Genes") +
  labs(x="Gene Ratio", y="GO Terms") +
  theme(legend.text = element_text(size = 12),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=12),
        legend.title = element_text(size=12,face="bold"), 
        axis.title=element_text(size=12, face="bold"))

dotplot(ego_genesDown) + ggtitle("GO Over-representation Downregulated Genes") +
  labs(x="Gene Ratio", y="GO Terms") +
  theme(legend.text = element_text(size = 12),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=12),
        legend.title = element_text(size=12,face="bold"), 
        axis.title=element_text(size=12, face="bold"))
# emapplot(ego_genesUp)
# emapplot(ego_genesDown)
# cnetplot(ego_genesUp, categorySize="pvalue", foldChange=geneList_up)
# cnetplot(ego_genesDown, categorySize="pvalue", foldChange=geneList_down)
# 
# 
# ego_genesUp_df <- as.data.frame(ego_genesUp) 
# sorted_egoUp <- ego_genesUp_df[order(-ego_genesUp_df$Count),]
# sorted_egoUp_top10 <- head(sorted_egoUp, 10)
# egoUp_top10_genes_all <- unlist(strsplit(sorted_egoUp_top10$geneID, "[/]"))
# egoUp_top10_genes_group <- strsplit(sorted_egoUp_top10$geneID, "[/]")
# egoUp_top10_genes_unique <- unique(egoUp_top10_genes)
# table(egoUp_top10_genes)
# egoUp_genesByGroup <- as.data.frame(t(plyr::ldply(egoUp_top10_genes_group, rbind)))
# colnames(egoUp_genesByGroup) <- sorted_egoUp_top10$Description
# egoUp_genesByGroup_ionOnly <- egoUp_genesByGroup[,c(1:6,8:10)]
# # write.csv(egoUp_genesByGroup, file="top10GOtermsUpregulated_geneMembership.csv")
# ionGenes <- unique(unlist(egoUp_genesByGroup_ionOnly))

# library(biomaRt)
# ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
# IDs <- as.character(ionGenes)
# geneUpID <- names(geneList_up)
# geneDownID <- names(geneList_down)
# genedesc_ion <- getBM(attributes=c('external_gene_name','description'), filters = 'external_gene_name', values = IDs, mart =ensembl)
# write.csv(genedesc_ion, file = "ionChannelGenes_description.csv")
# 
# genedesc_Up <- getBM(attributes=c('external_gene_name','description'), filters = 'external_gene_name', values = geneUpID, mart =ensembl)
# write.csv(genedesc_Up, file = "upregulatedGenes_description.csv")
# genedesc_Down <- getBM(attributes=c('external_gene_name','description'), filters = 'external_gene_name', values = geneDownID, mart =ensembl)
# write.csv(genedesc_Down, file = "downrgulatedGenes_description.csv")
```


```{r}
geneList_all <- as.vector(results_0to8d$log2FoldChange)
names(geneList_all) <- rownames(results_0to8d)
a <- names(geneList_all)
genes_ENTREZID <- bitr(a, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = OrgDB)$ENTREZID
names(geneList_all) <- genes_ENTREZID

gene_df <- data.frame(Entrez=names(geneList_all), FC=geneList_all)
gene_df <- gene_df[abs(gene_df$FC) > 1,]
gene_df$group <- "upregulated"
gene_df$group[gene_df$FC < 0] <- "downregulated"
gene_df$othergroup <- "A"
gene_df$othergroup[abs(gene_df$FC) > 2] <- "B"

formula_res <- compareCluster(Entrez~group+othergroup, data=gene_df, fun="enrichKEGG")
head(as.data.frame(formula_res))

```
#### 3. Exploring Results

```{r}
plotMA(res, ylim=c(-2,2))
plotCounts(dds, gene=which.min(res$padj), intgroup="treatment")

```

#### Log normalize results ####
```{r}

# normalizedCounts <- t( t(counts(dds)) / sizeFactors(dds) )

#log2 normalized counts
rld2 <- rlog(dds, blind = FALSE)
save(rld2, file = "RLD2_SC1,7,10_Timecourse_hmap.RData")

# load("RLD2_SC1,7,10_Timecourse_hmap.RData")
```

#### Clustering ###

```{r}
sampleDists <- dist(t(assay(rld2)))
sampleDists

library("pheatmap")
library("RColorBrewer")

sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste(rld2$treatment, rld2$cell, sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)

library("PoiClaClu")
poisd <- PoissonDistance(t(counts(dds)))
samplePoisDistMatrix <- as.matrix( poisd$dd )
rownames(samplePoisDistMatrix) <- paste( dds$dex, dds$cell, sep=" - " )
colnames(samplePoisDistMatrix) <- NULL
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         col = colors)

mds <- as.data.frame(colData(rld2))  %>%
         cbind(cmdscale(sampleDistMatrix))
ggplot(mds, aes(x = `1`, y = `2`, color = cell, shape = treatment)) +
  geom_point(size = 3) + coord_fixed() + theme_bw() +
  xlab("PC1") + ylab("PC2") +
  theme(legend.text = element_text(size = 10), 
        plot.title = element_text(size = 14, 
                                  hjust = 0.5, 
                                  face = "bold"), 
        axis.text=element_text(size=12),
        legend.title = element_text(size=12,face="bold"),
        # legend.position = "bottom",
        axis.title=element_text(size=12))

library("genefilter")
topVarGenes <- head(order(rowVars(assay(rld2)), decreasing = TRUE), 5000)
mat  <- assay(rld2)[ topVarGenes, ]
mat  <- mat - rowMeans(mat)
anno <- as.data.frame(colData(rld2)[, c("cell","treatment")])
names(anno) <- c("Cell", "Treatment")
annotation_colors = list(
  Cell = c(SC01="orange", SC07="cyan", SC10="purple"),
  Treatment = c("0"="grey90", "3"="grey60", "8"="grey30"))
pheatmap(mat, annotation_col = anno, show_rownames = F, show_colnames = F,
         annotation_colors = annotation_colors)
```

#### Time series analysis ####

# 1 DESeq2 time series analysis
```{r}
browseVignettes("rnaseqGene")

ddsTC <- DESeq(dds, test="LRT", reduced = ~ cell + treatment)
resTC <- results(ddsTC)
resTC$symbol <- mcols(ddsTC)$symbol
# head(resTC[order(resTC$padj),], 4)

tc <- plotCounts(ddsTC, which.min(resTC$padj), 
                   intgroup = c("treatment","cell"), returnData = TRUE)

ddsTC[which.min(resTC$padj),]

ggplot(tc,
  aes(x = rep(c(0,3,8), each=9), y = count, color = cell, group = cell)) + 
  geom_point() + geom_smooth(se = FALSE, method = "loess") + scale_y_log10() +
  theme_bw() +
  ggtitle("Time Course Expression of PDK4") +
  labs(x="Time (days)", y="Gene Count") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.text = element_text(size = 12),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=12),
        legend.title = element_text(size=12,face="bold"), 
        axis.title=element_text(size=12, face="bold"),
        legend.position = "bottom")

resultsNames(ddsTC)

betas <- coef(ddsTC)
colnames(betas)

topGenes <- head(order(resTC$padj),50)
mat <- betas[topGenes, -c(1,2)]
thr <- 3 
mat[mat < -thr] <- -thr
mat[mat > thr] <- thr
pheatmap(mat, breaks=seq(from=-thr, to=thr, length=101),
         cluster_col=FALSE)
```

```{r}
# 1.1 ANOVA - compare btwn sublines 
group<-as.factor(c(1,1,1,2,2,2,3,3,3))
# Getting anova values for each gene in dataset
anova_baseline <- list()
TukeySC07toSC01 <- list()
TukeySC10toSC01 <- list()
TukeySC10toSC07 <- list()
norm_data <- as.data.frame(assay(rld2))[c('SC01_day0_rep1',
                                        'SC01_day0_rep2',
                                        'SC01_day0_rep3',
                                        'SC07_day0_rep1',
                                        'SC07_day0_rep2',
                                        'SC07_day0_rep3',
                                        'SC10_day0_rep1',
                                        'SC10_day0_rep2',
                                        'SC10_day0_rep3')]
for (gene in 1:nrow(norm_data)) {
  gene_norm_data <- norm_data[gene,]
  # d3 <- data.frame(y = gene_norm_data, group = group)
  # fit <- lm(y~group, d3)
  gene_norm_data_melt <- melt(gene_norm_data)
  gene_norm_data_melt$group <- group
  fit <- aov(value~group, gene_norm_data_melt)
  # anova_list[gene] <- anova(fit)$'Pr(>F)'[1]
  anova_baseline[gene] <- summary(fit)[[1]][['Pr(>F)']][1]
  results <- TukeyHSD(fit, conf.level = 0.95)
  TukeySC07toSC01[gene] <- results$group[,'p adj'][1]
  TukeySC10toSC01[gene] <- results$group[,'p adj'][2]
  TukeySC10toSC07[gene] <- results$group[,'p adj'][3]
  
}
# print(anova_list)
anova_pval <- unlist(anova_baseline) # make array
TukeySC07toSC01_pval <- unlist(TukeySC07toSC01)
TukeySC10toSC01_pval <- unlist(TukeySC10toSC01)
TukeySC10toSC07_pval <- unlist(TukeySC10toSC07)

# Make master dataset with statistics
norm_data_stats <- as.data.frame(norm_data)
norm_data_stats <- cbind(norm_data_stats, anova_pval)
norm_data_stats <- cbind(norm_data_stats, TukeySC07toSC01_pval)
norm_data_stats <- cbind(norm_data_stats, TukeySC10toSC01_pval)
norm_data_stats <- cbind(norm_data_stats, TukeySC10toSC07_pval)

save(norm_data_stats, file = "subcloneCounts_anova_tukey_DESeq2.RData")

# Identify genes that differ between clones based on 
# ANOVA p-value with defined threshold
sigThresh <- 0.05
table(anova_pval < sigThresh)
table(TukeySC07toSC01_pval < sigThresh)
table(TukeySC10toSC01_pval < sigThresh)
table(TukeySC10toSC07_pval < sigThresh)

sigIndecies <- which(norm_data_stats["anova_pval"] < sigThresh)

sigIndeciesAll <- which(norm_data_stats["anova_pval"] < sigThresh & 
                          norm_data_stats["TukeySC07toSC01_pval"] < sigThresh &
                          norm_data_stats["TukeySC10toSC01_pval"] < sigThresh &
                          norm_data_stats["TukeySC10toSC07_pval"] < sigThresh)

sigDiffGenes <- rownames(norm_data_stats[sigIndecies,])
sigDiffGenesAll <- rownames(norm_data_stats[sigIndeciesAll,])
```

# 2. ANOVA btwn time points & shared btwn sublines)
```{r}
group<-as.factor(c(1,1,1,2,2,2,3,3,3))
# Getting anova values for each gene in dataset
anova_SC01 <- list()
TukeySC01_time0 <- list()
TukeySC01_time3 <- list()
TukeySC01_time8 <- list()
norm_data_SC01time <- as.data.frame(assay(rld2))[c('SC01_day0_rep1',
                                        'SC01_day0_rep2',
                                        'SC01_day0_rep3',
                                        'SC01_day3_rep1',
                                        'SC01_day3_rep2',
                                        'SC01_day3_rep3',
                                        'SC01_day8_rep1',
                                        'SC01_day8_rep2',
                                        'SC01_day8_rep3')]
for (gene in 1:nrow(norm_data_SC01time)) {
  gene_norm_data <- norm_data_SC01time[gene,]
  # d3 <- data.frame(y = gene_norm_data, group = group)
  # fit <- lm(y~group, d3)
  gene_norm_data_melt <- melt(gene_norm_data)
  gene_norm_data_melt$group <- group
  fit <- aov(value~group, gene_norm_data_melt)
  # anova_list[gene] <- anova(fit)$'Pr(>F)'[1]
  anova_SC01[gene] <- summary(fit)[[1]][['Pr(>F)']][1]
  results <- TukeyHSD(fit, conf.level = 0.95)
  TukeySC01_time0[gene] <- results$group[,'p adj'][1]
  TukeySC01_time3[gene] <- results$group[,'p adj'][2]
  TukeySC01_time8[gene] <- results$group[,'p adj'][3]
  
}
# print(anova_list)
anova_SC01_pval <- unlist(anova_SC01) # make array
TukeySC01_time0_pval <- unlist(TukeySC01_time0)
TukeySC01_time3_pval <- unlist(TukeySC01_time3)
TukeySC01_time8_pval <- unlist(TukeySC01_time8)

# Make master dataset with statistics
norm_data_stats_SC01 <- as.data.frame(norm_data_SC01time)
norm_data_stats_SC01 <- cbind(norm_data_stats_SC01, anova_SC01_pval)
norm_data_stats_SC01 <- cbind(norm_data_stats_SC01, TukeySC01_time0_pval)
norm_data_stats_SC01 <- cbind(norm_data_stats_SC01, TukeySC01_time3_pval)
norm_data_stats_SC01 <- cbind(norm_data_stats_SC01, TukeySC01_time8_pval)

save(norm_data_stats_SC01, file = "subcloneCounts_anova_tukey_DESeq2_SC01time.RData")

# Identify genes that differ between clones based on 
# ANOVA p-value with defined threshold
sigThresh <- 0.05
table(anova_SC01_pval < sigThresh)
table(TukeySC01_time0_pval < sigThresh)
table(TukeySC01_time3_pval < sigThresh)
table(TukeySC01_time8_pval < sigThresh)

sigIndecies_SC01 <- which(norm_data_stats_SC01["anova_SC01_pval"] < sigThresh)

sigIndeciesAll_SC01 <- which(norm_data_stats_SC01["anova_SC01_pval"] < sigThresh & 
                          norm_data_stats_SC01["TukeySC01_time0_pval"] < sigThresh &
                          norm_data_stats_SC01["TukeySC01_time3_pval"] < sigThresh &
                          norm_data_stats_SC01["TukeySC01_time8_pval"] < sigThresh)

sigDiffGenes_SC01 <- rownames(norm_data_stats_SC01[sigIndecies_SC01,])
sigDiffGenesAll_SC01 <- rownames(norm_data_stats_SC01[sigIndeciesAll_SC01,])
```

```{r}
group<-as.factor(c(1,1,1,2,2,2,3,3,3))
# Getting anova values for each gene in dataset
anova_SC07 <- list()
TukeySC07_time0 <- list()
TukeySC07_time3 <- list()
TukeySC07_time8 <- list()
norm_data_SC07time <- as.data.frame(assay(rld2))[c('SC07_day0_rep1',
                                        'SC07_day0_rep2',
                                        'SC07_day0_rep3',
                                        'SC07_day3_rep1',
                                        'SC07_day3_rep2',
                                        'SC07_day3_rep3',
                                        'SC07_day8_rep1',
                                        'SC07_day8_rep2',
                                        'SC07_day8_rep3')]
for (gene in 1:nrow(norm_data_SC07time)) {
  gene_norm_data <- norm_data_SC07time[gene,]
  # d3 <- data.frame(y = gene_norm_data, group = group)
  # fit <- lm(y~group, d3)
  gene_norm_data_melt <- melt(gene_norm_data)
  gene_norm_data_melt$group <- group
  fit <- aov(value~group, gene_norm_data_melt)
  # anova_list[gene] <- anova(fit)$'Pr(>F)'[1]
  anova_SC07[gene] <- summary(fit)[[1]][['Pr(>F)']][1]
  results <- TukeyHSD(fit, conf.level = 0.95)
  TukeySC07_time0[gene] <- results$group[,'p adj'][1]
  TukeySC07_time3[gene] <- results$group[,'p adj'][2]
  TukeySC07_time8[gene] <- results$group[,'p adj'][3]
  
}
# print(anova_list)
anova_SC07_pval <- unlist(anova_SC07) # make array
TukeySC07_time0_pval <- unlist(TukeySC07_time0)
TukeySC07_time3_pval <- unlist(TukeySC07_time3)
TukeySC07_time8_pval <- unlist(TukeySC07_time8)

# Make master dataset with statistics
norm_data_stats_SC07 <- as.data.frame(norm_data_SC07time)
norm_data_stats_SC07 <- cbind(norm_data_stats_SC07, anova_SC07_pval)
norm_data_stats_SC07 <- cbind(norm_data_stats_SC07, TukeySC07_time0_pval)
norm_data_stats_SC07 <- cbind(norm_data_stats_SC07, TukeySC07_time3_pval)
norm_data_stats_SC07 <- cbind(norm_data_stats_SC07, TukeySC07_time8_pval)

save(norm_data_stats_SC07, file = "subcloneCounts_anova_tukey_DESeq2_SC07time.RData")

# Identify genes that differ between clones based on 
# ANOVA p-value with defined threshold
sigThresh <- 0.05
table(anova_SC07_pval < sigThresh)
table(TukeySC07_time0_pval < sigThresh)
table(TukeySC07_time3_pval < sigThresh)
table(TukeySC07_time8_pval < sigThresh)

sigIndecies_SC07 <- which(norm_data_stats_SC07["anova_SC07_pval"] < sigThresh)

sigIndeciesAll_SC07 <- which(norm_data_stats_SC07["anova_SC07_pval"] < sigThresh & 
                          norm_data_stats_SC07["TukeySC07_time0_pval"] < sigThresh &
                          norm_data_stats_SC07["TukeySC07_time3_pval"] < sigThresh &
                          norm_data_stats_SC07["TukeySC07_time8_pval"] < sigThresh)

sigDiffGenes_SC07 <- rownames(norm_data_stats_SC07[sigIndecies_SC07,])
sigDiffGenesAll_SC07 <- rownames(norm_data_stats_SC07[sigIndeciesAll_SC07,])
```

```{r}
group<-as.factor(c(1,1,1,2,2,2,3,3,3))
# Getting anova values for each gene in dataset
anova_SC10 <- list()
TukeySC10_time0 <- list()
TukeySC10_time3 <- list()
TukeySC10_time8 <- list()
norm_data_SC10time <- as.data.frame(assay(rld2))[c('SC10_day0_rep1',
                                        'SC10_day0_rep2',
                                        'SC10_day0_rep3',
                                        'SC10_day3_rep1',
                                        'SC10_day3_rep2',
                                        'SC10_day3_rep3',
                                        'SC10_day8_rep1',
                                        'SC10_day8_rep2',
                                        'SC10_day8_rep3')]
for (gene in 1:nrow(norm_data_SC10time)) {
  gene_norm_data <- norm_data_SC10time[gene,]
  # d3 <- data.frame(y = gene_norm_data, group = group)
  # fit <- lm(y~group, d3)
  gene_norm_data_melt <- melt(gene_norm_data)
  gene_norm_data_melt$group <- group
  fit <- aov(value~group, gene_norm_data_melt)
  # anova_list[gene] <- anova(fit)$'Pr(>F)'[1]
  anova_SC10[gene] <- summary(fit)[[1]][['Pr(>F)']][1]
  results <- TukeyHSD(fit, conf.level = 0.95)
  TukeySC10_time0[gene] <- results$group[,'p adj'][1]
  TukeySC10_time3[gene] <- results$group[,'p adj'][2]
  TukeySC10_time8[gene] <- results$group[,'p adj'][3]
  
}
# print(anova_list)
anova_SC10_pval <- unlist(anova_SC10) # make array
TukeySC10_time0_pval <- unlist(TukeySC10_time0)
TukeySC10_time3_pval <- unlist(TukeySC10_time3)
TukeySC10_time8_pval <- unlist(TukeySC10_time8)

# Make master dataset with statistics
norm_data_stats_SC10 <- as.data.frame(norm_data_SC10time)
norm_data_stats_SC10 <- cbind(norm_data_stats_SC10, anova_SC10_pval)
norm_data_stats_SC10 <- cbind(norm_data_stats_SC10, TukeySC10_time0_pval)
norm_data_stats_SC10 <- cbind(norm_data_stats_SC10, TukeySC10_time3_pval)
norm_data_stats_SC10 <- cbind(norm_data_stats_SC10, TukeySC10_time8_pval)

save(norm_data_stats_SC10, file = "subcloneCounts_anova_tukey_DESeq2_SC10time.RData")

# Identify genes that differ between clones based on 
# ANOVA p-value with defined threshold
sigThresh <- 0.05
table(anova_SC10_pval < sigThresh)
table(TukeySC10_time0_pval < sigThresh)
table(TukeySC10_time3_pval < sigThresh)
table(TukeySC10_time8_pval < sigThresh)

sigIndecies_SC10 <- which(norm_data_stats_SC10["anova_SC10_pval"] < sigThresh)

sigIndeciesAll_SC10 <- which(norm_data_stats_SC10["anova_SC10_pval"] < sigThresh & 
                          norm_data_stats_SC10["TukeySC10_time0_pval"] < sigThresh &
                          norm_data_stats_SC10["TukeySC10_time3_pval"] < sigThresh &
                          norm_data_stats_SC10["TukeySC10_time8_pval"] < sigThresh)

sigDiffGenes_SC10 <- rownames(norm_data_stats_SC10[sigIndecies_SC10,])
sigDiffGenesAll_SC10 <- rownames(norm_data_stats_SC10[sigIndeciesAll_SC10,])
```

```{r}

all_SCs_time <- Reduce(intersect, 
                       list(sigDiffGenesAll_SC01,
                            sigDiffGenesAll_SC07,
                            sigDiffGenesAll_SC10))

df_allSCs_time <- data.frame(gene = all_SCs_time)
library('biomaRt')
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
genes <- df_allSCs_time$gene
G_list <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","hgnc_symbol"),values=genes,mart=mart)
merge(df_allSCs_time,G_list,by.x="gene",by.y="ensembl_gene_id")

write.csv(G_list, file = "ANOVA_allSCsTime_shared_genes.csv")

# Compare gene lists for between sublines and time
library(devtools)
# install_github("wjawaid/enrichR")
library(enrichR)
dbs <- listEnrichrDbs()
head(dbs)
dbs <- c("KEGG_2016", 
         "GO_Molecular_Function_2018")

enriched_allSCstime <- enrichr(G_list$hgnc_symbol, dbs)
library(ggplot2)

KEGG_upreg_allSCstime_top5 <- enriched_allSCstime[["KEGG_2016"]][1:5,]
KEGG_upreg_allSCstime_top5$Terms <- factor(KEGG_upreg_allSCstime_top5$Term,
                                           levels=KEGG_upreg_allSCstime_top5$Term)

ggplot(KEGG_upreg_allSCstime_top5, 
       aes(x=Terms, y=-log10(Adjusted.P.value))) +
  geom_bar(stat='identity') +
  coord_flip() +
  labs(x = "Pathway Term", y = "-log10(q-value)")  +
  theme_bw()  + ggtitle("Pathway Enrichment - KEGG 2016") +
  theme(plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=12), 
        axis.title=element_text(size=14,face="bold"))
  

```

# 3 Jack's method 
```{r}
#Grab all the names from res in the DESeq matrix
topGenes <- which(res$padj <= 0.001)

# countMAT = data.frame(normalizedCounts[topGenes,])

subrl = data.frame(assay(rld2))
rlMAT = data.frame(subrl[topGenes,])

#Labeling rows with ENSG IDs
# countMAT$ensembl_gene_id = row.names(countMAT)
# countMAT$padj = res[topGenes,"padj"]

rlMAT$ensembl_gene_id = row.names(rlMAT)
rlMAT$padj = res[topGenes,"padj"]

library(biomaRt)
ensembl <- useMart("ensembl")
mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
genes = row.names(rlMAT)
G_list <- getBM(attributes= c("ensembl_gene_id","hgnc_symbol"),
                filters= "ensembl_gene_id",
                values=genes,
                mart=mart)

#Check if data fits a normal distribution
# plot(density(c(as.matrix(countMAT[,1:27]))))
plot(density(c(as.matrix(rlMAT[,1:27]))))


#rlMAT follows a normal distribution, therefore we will use this in the heatmap construction
#Labeling df with hgnc symbols
GE_data <- merge(G_list, rlMAT, by = "ensembl_gene_id")

#Making rownames unique hgnc symbols
rownames(GE_data) <- make.names(GE_data[,"hgnc_symbol"], unique = TRUE)
GE_data = GE_data[order(GE_data$padj),]


#Averaging rld between trials
Acol <- c("SC01_day0",
          "SC01_day3",
          "SC01_day8",
          "SC07_day0",
          "SC07_day3",
          "SC07_day8",
          "SC10_day0",
          "SC10_day3",
          "SC10_day8")
for(i in 1:length(Acol)){
  j = 2+i
  k = 2+3*i
  GE_data[,Acol[i]] = rowMeans(GE_data[,c(j:k)])
}


#Calculating fold changes across conditions in a triangular matrix form
GE_mean = GE_data[,c(1,2,30:39)]
DEProc = GE_mean
startcol = 4
endcol = 12

allFC <- function(DEProc,startcol,endcol){ 
  GE_fold = DEProc[,-c(startcol:endcol)]
  colvec = colnames(DEProc)[startcol:endcol]
  
  #Last index is a self comparison and is removed
  for(k in 1:(length(colvec)-1)){
    #Start with column that is 1 away from index 
    for(j in (k+1):length(colvec)){
      compnam = paste0(colvec[j],"/",colvec[k])
      #Loop through each gene/row  
      for(i in 1:nrow(DEProc)){
        f = DEProc[i,colvec[j]]
        h = DEProc[i,colvec[k]]
        #Capture upregulation and down regulation
        if(f>h){
          # GE_fold[i,compnam] = 2^(f-h)
          GE_fold[i,compnam] = log2(f/h)
        }else{
          # GE_fold[i,compnam] = -2^(h-f)
          GE_fold[i,compnam] = log2(f/h)
        }
        
      }
    }
  }
  
  return(GE_fold)
  
}

GE_fold <- allFC(DEProc, 4, 12)

#Subset gene, then plot, then save plot
#Perhaps make heatmaps with scaled z scores
#Is there a way to consolidate replicate z scores? Geometric mean? 
#Regular mean, then scale.

# ImpRat = colnames(GE_fold)[c(4,5,6,9,12,14,17,21,24,25,26,27,30,32,36,37,38,39)]

#Listing of all important comparisons?
ImpRat = c("SC01_day3/SC01_day0", "SC01_day8/SC01_day3", "SC01_day8/SC01_day0", 
           "SC07_day3/SC07_day0", "SC07_day8/SC07_day3", "SC07_day8/SC07_day0", 
           "SC10_day3/SC10_day0", "SC10_day8/SC10_day3", "SC10_day8/SC10_day0", 
           "SC07_day0/SC01_day0", "SC10_day0/SC01_day0", "SC10_day0/SC07_day0",
           "SC07_day3/SC01_day3", "SC10_day3/SC01_day3", "SC10_day3/SC07_day3",
           "SC07_day8/SC01_day8", "SC10_day8/SC01_day8", "SC10_day8/SC07_day8" )
Imp_fold = GE_fold[,c("ensembl_gene_id", "hgnc_symbol", "padj", ImpRat)]
Imp_fold2 = Imp_fold[rowSums(abs(Imp_fold[,4:21])>=1.5)>=1,]

write.table(Imp_fold,"SC1,7,10-TimecoursePLX-ImportantFC_20180722.txt", sep="\t", row.names=F)

Imp_fold = read.delim("SC1,7,10-TimecoursePLX-ImportantFC_20180722.txt", sep="\t")

#Subset the LF mean of important genes from Log2 Fold Change (LFC) comparison data frame.
GE_Imp = subset(GE_mean,GE_mean$ensembl_gene_id%in%Imp_fold2$ensembl_gene_id)

library(pheatmap)

Necro = read.delim("KEGGNecroptosis_hsa04217_06-25-18.txt", header=T, stringsAsFactors = F)
Necro = Necro[rowSums(is.na(Necro)) == 0, ]
DE_Necro = merge(GE_Imp, Necro, by.x = "hgnc_symbol", by.y = "GeneName")
row.names(DE_Necro) = make.names(DE_Necro[,"hgnc_symbol"], unique = TRUE)
pheatmap(DE_Necro[3:29],cluster_cols = TRUE)
write.table(DE_Necro, "KEGGNecroptosis SC1,7,10 DESeq LRT.txt", sep="\t", row.names=FALSE, quote=FALSE)


Apop = read.delim("KEGGApoptosis_hsa04210_06-25-18.txt", header=T, stringsAsFactors = F)
Apop = Apop[rowSums(is.na(Apop)) == 0, ]
DE_Apop = merge(GE_Imp, Apop, by.x = "hgnc_symbol", by.y = "GeneName")
row.names(DE_Apop) = make.names(DE_Apop[,"hgnc_symbol"], unique = TRUE)
pheatmap(DE_Apop[3:29],cluster_cols = TRUE, scale = "row")
write.table(DE_Apop, "KEGGApoptosis SC1,7,10 DESeq LRT.txt", sep="\t", row.names=FALSE, quote=FALSE)

Ferr = read.delim("~/Documents/QuarantaLab/Ferroptosis/RNAseq/KEGGFerroptosis_hsa04216_06-25-18.txt", header=T, stringsAsFactors = F)
Ferr = Ferr[rowSums(is.na(Ferr)) == 0, ]
DE_Ferr = merge(GE_Imp, Ferr, by.x = "hgnc_symbol", by.y = "GeneName")
row.names(DE_Ferr) = make.names(DE_Ferr[,"hgnc_symbol"], unique = TRUE)
pheatmap(DE_Ferr[4:12],cluster_cols=FALSE, scale = "row")
write.table(DE_Ferr, "KEGGFerroptosis SC1,7,10 DESeq LRT.txt", sep="\t", row.names=FALSE, quote=FALSE)


```

#### 4. Different LC comparisons. Between subclones and at baseline vs idling.
Zscore heatmaps of relevant comparisons can be made as in above to visualize.

```{r}

#USES ABOVE CODE TO LINE 280. Run that pseudo-function.

library(pheatmap)
#Comparisons of difEx between subclones at baseline and idling
BetweenBase = c("SC07_day0/SC01_day0", "SC10_day0/SC01_day0", "SC10_day0/SC07_day0")
BetweenIdle = c("SC07_day8/SC01_day8", "SC10_day8/SC01_day8", "SC10_day8/SC07_day8")


#Unsure of how strict to make the cutoff. Should all the genes between clones be differentially expressed (3) or is a single difference sufficient?
Btw_b = GE_fold[,c("ensembl_gene_id", "hgnc_symbol", "padj", BetweenBase)]
Btw_b1 = Btw_b[rowSums(abs(Btw_b[,4:6])>=1.5)>=1,]
Btw_b2 = Btw_b[rowSums(abs(Btw_b[,4:6])>=1.5)>=2,]
Btw_b3 = Btw_b[rowSums(abs(Btw_b[,4:6])>=1.5)>=3,]

Btw_i = GE_fold[,c("ensembl_gene_id", "hgnc_symbol", "padj", BetweenIdle)]
Btw_i1 = Btw_i[rowSums(abs(Btw_i[,4:6])>=1.5)>=1,]
Btw_i2 = Btw_i[rowSums(abs(Btw_i[,4:6])>=1.5)>=2,]
Btw_i3 = Btw_i[rowSums(abs(Btw_i[,4:6])>=1.5)>=3,]

#This does not account for same direction of change. This can be plotted in a heatmap to view.
#Members that were "lost" by the baseline condition at being different. Were no longer found diffEx between the subclones when comparing baseline to idling DEGs.
LostDEG_b_1 = subset(Btw_b1,!Btw_b1$ensembl_gene_id%in%Btw_i1$ensembl_gene_id)
LostDEG_b_2 = subset(Btw_b2,!Btw_b2$ensembl_gene_id%in%Btw_i2$ensembl_gene_id)
LostDEG_b_3 = subset(Btw_b3, !Btw_b3$ensembl_gene_id%in%Btw_i3$ensembl_gene_id)

##Make heatmap
LostDEG_b_3_mean = subset(GE_mean,GE_mean$ensembl_gene_id%in%LostDEG_b_3$ensembl_gene_id)
row.names(LostDEG_b_3_mean) = make.names(LostDEG_b_3_mean[,"hgnc_symbol"], unique = TRUE)
pheatmap(LostDEG_b_3_mean[4:12],cluster_cols=FALSE, scale = "row")

#Members that remained different. 
StaticDEG_b_1 = subset(Btw_b1,Btw_b1$ensembl_gene_id%in%Btw_i1$ensembl_gene_id)
StaticDEG_b_2 = subset(Btw_b2,Btw_b2$ensembl_gene_id%in%Btw_i2$ensembl_gene_id)
StaticDEG_b_3 = subset(Btw_b3, Btw_b3$ensembl_gene_id%in%Btw_i3$ensembl_gene_id)

##Some HGNC_symbols are duplicates! I switched to ensembl_gene_id to fix.
StaticDEG_i_3 = subset(Btw_i3, Btw_i3$ensembl_gene_id%in%Btw_b3$ensembl_gene_id)


##Make heatmap
StaticDEG_b_3_mean = subset(GE_mean,GE_mean$ensembl_gene_id%in%StaticDEG_b_3$ensembl_gene_id)
row.names(StaticDEG_b_3_mean) = make.names(StaticDEG_b_3_mean[,"hgnc_symbol"], unique = TRUE)
pheatmap(StaticDEG_b_3_mean[4:12],cluster_cols=FALSE, scale = "row")


#Members that "gained" differences between the subclones in idling.  
GainDEG_i_1 = subset(Btw_i1, !Btw_i1$ensembl_gene_id%in%Btw_b1$ensembl_gene_id)
GainDEG_i_2 = subset(Btw_i2, !Btw_i2$ensembl_gene_id%in%Btw_b2$ensembl_gene_id)
GainDEG_i_3 = subset(Btw_i3, !Btw_i3$ensembl_gene_id%in%Btw_b3$ensembl_gene_id)

##Make heatmap
GainDEG_i_3_mean = subset(GE_mean,GE_mean$ensembl_gene_id%in%GainDEG_i_3$ensembl_gene_id)
row.names(GainDEG_i_3_mean) = make.names(GainDEG_i_3_mean[,"hgnc_symbol"], unique = TRUE)
pheatmap(GainDEG_i_3_mean[4:12],cluster_cols=FALSE, scale = "row")


#Members that were differentially expressed between idling (8day) and baseline within subclones. Those with shared diffEx may be convergent across multiple subclones depending on direction of expresison change.
Endpoint = c("SC01_day8/SC01_day0", "SC07_day8/SC07_day0", "SC10_day8/SC10_day0")
BtoIdle = GE_fold[,c("ensembl_gene_id", "hgnc_symbol", "padj", Endpoint)]
BtoIdle_1 = BtoIdle[rowSums(abs(BtoIdle[,4:6])>=1.5)>=1,]
BtoIdle_2 = BtoIdle[rowSums(abs(BtoIdle[,4:6])>=1.5)>=2,]
BtoIdle_3 = BtoIdle[rowSums(abs(BtoIdle[,4:6])>=1.5)>=3,]

##Make heatmap
BtoIdle_2_mean = subset(GE_mean,GE_mean$ensembl_gene_id%in%BtoIdle_2$ensembl_gene_id)
row.names(BtoIdle_2_mean) = make.names(BtoIdle_2_mean[,"hgnc_symbol"], unique = TRUE)

BtoIdle_2_mean_incExp = BtoIdle_2_mean[which(BtoIdle_2_mean$SC01_day0 < BtoIdle_2_mean$SC01_day8),]
BtoIdle_2_mean_incExp = BtoIdle_2_mean_incExp[which(BtoIdle_2_mean_incExp$SC07_day0 < BtoIdle_2_mean_incExp$SC07_day8),]
BtoIdle_2_mean_incExp[which(BtoIdle_2_mean_incExp$SC10_day0 < BtoIdle_2_mean_incExp$SC10_day8),]

LostDEG_b_2_mean = subset(GE_mean,GE_mean$ensembl_gene_id%in%LostDEG_b_2$ensembl_gene_id)
row.names(LostDEG_b_2_mean) = make.names(LostDEG_b_2_mean[,"hgnc_symbol"], unique = TRUE)
pheatmap(LostDEG_b_2_mean[4:12],cluster_cols=FALSE, scale = "row")

BtoIdleIncExp_DEbetweenSCs = BtoIdle_2_mean_incExp[which(row.names(BtoIdle_2_mean_incExp) %in% row.names(LostDEG_b_2_mean)),]

pheatmap(BtoIdle_2_mean_incExp[4:12],cluster_cols=FALSE, scale = "row")

library(devtools)
# install_github("wjawaid/enrichR")
library(enrichR)
dbs <- listEnrichrDbs()
head(dbs)
dbs <- c("GO_Molecular_Function_2015", "GO_Cellular_Component_2015", "GO_Biological_Process_2015")

enriched <- enrichr(row.names(BtoIdle_2_mean_incExp), dbs)

View(enriched[["GO_Molecular_Function_2015"]])
View(enriched[["GO_Cellular_Component_2015"]])
View(enriched[["GO_Biological_Process_2015"]])

enriched_MF_sig <- enriched[["GO_Molecular_Function_2015"]][enriched[["GO_Molecular_Function_2015"]]$Adjusted.P.value<0.05,]
enriched_MF_sig_df <- data.frame(enriched_MF_sig[,c(1,4,9)])
write.csv(enriched_MF_sig_df, "enriched_MF_significant.csv")

enriched_BP_sig <- enriched[["GO_Biological_Process_2015"]][enriched[["GO_Biological_Process_2015"]]$Adjusted.P.value<0.05,]
enriched_BP_sig_df <- data.frame(enriched_BP_sig[,c(1,4,9)])
write.csv(enriched_BP_sig_df, "enriched_BP_significant.csv")

Gini_scGenes <- c("APOE", "BCAN", "CES1", "CITED1",
                  "CPM", "CTSF", "DCT", "EDNRB", 
                  "EGR1", "ESRP1", "FSTL1", "MALAT1",
                  "MAP2K6", "MCF2L", "MLANA", "MXD4",
                  "OCA2", "PMEL", "SEMA6A", "SNAI2",
                  "SOX4", "TSPAN10")
enriched_sc <- enrichr(Gini_scGenes, dbs)

row.names(BtoIdle_2_mean_incExp) %in% Gini_scGenes

```



#### Rest of Jack's Analysis ####
```{r}
#Visually inspect trending members from heatmaps.
#Plots of specific trending members?
p <- ggplot(data=df2, aes(x=dose, y=len, fill=supp)) +
  geom_bar(stat="identity", color="black", position=position_dodge())+
  theme_minimal()
```


```{r}
#GLM Coef Heatmap.
betas <- coef(dds)
topGenes <- which(res$padj <= 0.001)
mat <- data.frame(betas[topGenes,])
mat$ensembl_gene_id = row.names(mat)
mat$padj = res[topGenes,"padj"]
ensembl <- useMart("ensembl")
mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
genes = row.names(mat)
G_list <- getBM(attributes= c("ensembl_gene_id","hgnc_symbol"),
                filters= "ensembl_gene_id",
                values=genes,
                mart=mart)

GE_data <- merge(mat, G_list, by = "ensembl_gene_id")
rownames(GE_data) <- make.names(GE_data[,"hgnc_symbol"], unique = TRUE)
GE_data = GE_data[order(GE_data$padj),]


#Sorting script to pick out entries greater than or less than +-1
eg = c()
for(i in 3:10){
  g = which(GE_data[,i] > 3 | GE_data[,i] < -3)
  eg = c(eg,g)
}
eg = unique(eg)

mat = GE_data[eg,-c(1:2,11,12)]
thr <- 3 
mat[mat < -thr] <- -thr
mat[mat > thr] <- thr
library(pheatmap)
pheatmap(mat, cluster_cols = FALSE)

# ssdg = sdg[1:1000, ]
dim(sdg)
head(sdg)

```


